Quantum Hoare Type Theory

### 2020-11-06 ms-paper/ms-paper.tex
Remove QIO from bg, move HTT before QC, fix von Neumann name

This paper is organized as follows. We discuss background in the next section, \cref{sec:bg}; specifically, quantum computing (\cref{sec:qc}),  Hoare Type Theory in the classical setting (\cref{sec:htt}) and then the Quantum IO Monad interface (\cref{sec:qio}). Then we introduce our contributions in merging these ideas together for specification and enforcement of useful semantic properties in section~\cref{sec:qhtt} along with its syntax (\cref{sec:syntax}), typing rules (\cref{sec:typing}) and, some examples (\cref{sec:examples}). Some related work follows in \cref{sec:related}. Finally, we conclude and share ideas for future work in the final section (\cref{sec:conclusion}).

\section{Background}
\label{sec:bg}
In this section, we provide background on the core ideas from existing literature that form the foundation for our work --- Hoare Type Theory (HTT) and the Quantum IO Monad interface. We start with some basics of quantum computing.
% and refer the reader to Nielsen's excellent series of essays~\cite{matuschak2019} for a first introduction and to the standard textbook~\cite{nielsen2010} for advanced material.

@@

\begin{definition}[Hermitian operator]
	A Hermitian operator, $A$, in a finite-dimensional complex Hilbert space, $\mathcal{H}$, is a self-adjoint operator in $\mathcal{H}$. It is equal to its complex conjugate: $A = A^\dagger$.
\end{definition}
Hermitian operators describe physical observables and their eigenvalues are real numbers. \citet{dhondt_quantum_2006} define a quantum predicate in $\mathcal{H}$ as a Hermitian operator with eigenvalues within the unit interval ($[0, 1]$).

\citet{ying_duan_feng_ji_2009} point out a problem with the above approach that it is not always possible to define certain logical operations (such as conjunction and disjunction) on some of these quantum predicates as they do not for a lattice. These operations are necessary to be able to compositionally reason about quantum programs. They, in turn, restrict themselves to quantum predicates that are expressed using a special class of Hermitian operators, called projection operators or projectors. It is well-known that projection operators in Hilbert space correspond one to one with closed subspaces in that space. Further, \citet{logicqm36} showed that the set of closed subspaces of Hilbert space form an orthomodular lattice that they use to define a logic for quantum mechanics. Similar to \citet{ying_duan_feng_ji_2009}, we will restrict ourselves to projection operators (or equivalently closed subspaces) to define quantum predicates.

@@

\subsection{Quantum IO Monad}
\label{sec:qio}

The Quantum IO (QIO) monad~\cite{qio} is a purely functional interface for quantum programming that provides a separation between the unitary (reversible) and non-unitary (irreversible) fragments of quantum computation. It provides isolation of quantum effects inside a monad similar to what we need to provide in our Quantum Hoare monad.

QIO interface was developed as a library for the Haskell programming language but its design was influenced~\cite{asg2010} by the category \textbf{FQC} of finite quantum computations (where computations are interpreted as superoperators) as explored by \citet{fqc06} in previous work. This is the most we will say about the relation of QIO monad to category theory in this paper.

Here we discuss the relevant concepts from the QIO interface. QIO provides a type for referring to qubits, \texttt{Qbit}; a type for unitary operations, \texttt{U}; and, the \texttt{QIO} type operator which is a monad indexed by the result of the quantum computation. The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
mkQbit :: Bool -> QIO Qbit       -- initialization
applyU :: U -> QIO ()            -- apply a unitary to quantum state
measQbit :: Qbit -> QIO Bool     -- measurement
\end{lstlisting}
\end{minipage}

Further, \texttt{U} is defined as a monoid with neutral element, \texttt{mempty}, and operation, \texttt{mappend}, that encodes reversible operations on quantum state. The core unitary primitives that we will need are:

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rot :: Qbit -> ((Bool, Bool) -> C) -> U
cond :: Qbit -> (Bool -> U) -> U
\end{lstlisting}
\end{minipage}

\texttt{rot} takes a qubit and a two-by-two complex-valued unitary matrix (represented as a function from the matrix indices to $\mathbb{C}$) and lets one define arbitrary rotation on a single qubit. \texttt{rot} can then be used to define standard gates such as the Hadamard, H, and the Pauli gates, X, Y and Z. \texttt{cond} is the unitary conditional that is used to perform branching unitary operations. For the present paper, we use a simpler conditional, \texttt{ifQ}, that is defined as follows:

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
ifQ :: Qbit -> U -> U
ifQ q u = cond q (\x -> if x then u else mempty)
\end{lstlisting}
\end{minipage}

That is, \texttt{ifQ} acts as the standard control operator in the quantum circuit model that runs its second argument based on the truth value of its first argument. Then, it is easy to define controlled gates such as CNOT with the expression $\kw{ifQ q1 (X q2)}$.

Quantum state in the QIO interface is modeled as a normalized vector that stores pairs representing complex amplitudes associated with each basis state (represented as a map from Qbit to Bool types). The operations defined over the vector class ensure that the quantum state is kept normalized throughout.

QIO further ensures that the monoidal structure of unitary operations lets one run a unitary operation over the complete quantum state. We elide details here and refer the reader to \cite{qio} for more.

@@

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Propositions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{class}(\textbf{X}) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \kw{\textbf{init}} N \mid \kw{\textbf{apply}} U \kw{\textbf{to}} N \mid \kw{\textbf{meas}} N$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x = N; E $\\ \\
%		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

>>>>

%This paper is organized as follows. We discuss background in the next section, \cref{sec:bg}; specifically, quantum computing (\cref{sec:qc}), Hoare Type Theory in the classical setting (\cref{sec:htt}) and then the Quantum IO Monad interface (\cref{sec:qio}). Then we introduce our contributions in merging these ideas together for specification and enforcement of useful semantic properties in section~\cref{sec:qhtt} along with its syntax (\cref{sec:syntax}), typing rules (\cref{sec:typing}) and, some examples (\cref{sec:examples}). Some related work follows in \cref{sec:related}. Finally, we conclude and share ideas for future work in the final section (\cref{sec:conclusion}).

\section{Background}
\label{sec:bg}
In this section, we provide background on the core ideas from existing literature that form the foundation for our work --- Hoare Type Theory (HTT), quantum computing, and quantum logic defined using closed subspaces.

@@

\begin{definition}[Hermitian operator]
	A Hermitian operator, $A$, in a finite-dimensional complex Hilbert space, $\mathcal{H}$, is a self-adjoint operator in $\mathcal{H}$. It is equal to its complex conjugate: $A = A^\dagger$.
\end{definition}
%Hermitian operators describe physical observables and their eigenvalues are real numbers. \citet{dhondt_quantum_2006} define a quantum predicate in $\mathcal{H}$ as a Hermitian operator with eigenvalues within the unit interval ($[0, 1]$).

\citet{ying_duan_feng_ji_2009} restrict themselves to quantum predicates that are expressed using a special class of Hermitian operators, called projection operators or projectors. It is well-known that projection operators in Hilbert space correspond one to one with closed subspaces in that space. Further, \citet{logicqm36} showed that the set of closed subspaces of Hilbert space form an orthomodular lattice that they use to define a logic for quantum mechanics. Similar to \citet{ying_duan_feng_ji_2009}, we will restrict ourselves to projection operators (or equivalently closed subspaces) to define quantum predicates.

@@

@@

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Propositions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{class}(\textbf{X}) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \kw{\textbf{init}} N \mid \kw{\textbf{apply}} U \kw{\textbf{to}} N \mid \kw{\textbf{meas}} N$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x = N; E $\\ \\
		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}
